---
title: "Course Project 1"
author: "Charles Vuono"
date: "June 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Introduction

I did some stuff for this project

###Loading and preprocessing the data

We load the data from the file *activity.csv* using the R code:

```{r}
df<-read.csv("activity.csv", header=TRUE)
```

The data contains many NA values in the field *steps* which we will address later in this report. 

###What is mean total number of steps taken per day?

First we transform the data to find the total number of steps per day (leaving any day with a NA observation as NA so as not to skew the data). We accomplish this with the R code:

```{r}
dailyDF <- aggregate(df$steps, list(date=df$date), sum)
colnames(dailyDF)<-c("date", "daily.steps")
```

Before calculating the mean and median of steps per day, we can investigate the distribution of steps per day by considering the frequency histogram of the daily steps data.

```{r}
hist(dailyDF$daily.steps, col="pink", xlab="Steps per Day", main="Frequency Histogram of Steps per Day")
```

To find the mean and median steps per day, we must ignore the NA values in the data and calculate the mean and median as:

```{r}
mean(dailyDF$daily.steps, na.rm=TRUE)
median(dailyDF$daily.steps, na.rm=TRUE)
```



